{\Huge Supersymmetry} \\ \vspace{5mm} Lecture notes for FYS5190/FYS9190 Supersymmetry

### 2023-12-04 groups.tex
Fix reference

We can construct unitary matrices from either Hermitian or anti-Hermitian matrices by using the {\bf matrix exponential}. This formally interprets the exponential series in terms of a real or complex valued $n\times n$ matrix $M$ as
\[ \exp(M)\equiv\sum_{n=0}^\infty \frac{M^n}{n!}=I+M+\frac{1}{2}M^2+\frac{1}{6}M^3+\ldots. \]
This series can be shown to always converge, so the series is well defined. Since this is a series of non-commuting objects we again have to be careful with using properties of the exponential from ordinary arithmetic. The following useful properties of the matrix exponential can be proven:
\begin{itemize}
\item[i)] $\exp(A^T)=\exp(A)^T$
\item[ii)] $\exp(A^*)=\exp(A)^*$
\item[iii)] If $B$ is an invertible matrix then $\exp(BAB^{-1})=B\exp(A)B^{-1}$
\item[iv)] If $[A,B]=0$ then $\exp(A)\exp(B)=\exp(A+B)$
\item[v)] $\det e^A = e^{\Tr A}$. 
\end{itemize}

@@

One particular explicit form, where the angles correspond to the three {\bf Euler angles} of rotation in three dimensions, $\alpha$, $\beta$ and $\gamma$, is
\[
R(\alpha,\beta,\gamma)=\left[\begin{matrix} 
\cos\alpha\cos\beta\cos\gamma-\sin\alpha\sin\gamma & -\cos\alpha\cos\beta\sin\gamma-\sin\alpha\cos\gamma & \cos\alpha\sin\beta  \\ 
\sin\alpha\cos\beta\cos\gamma+\cos\alpha\sin\gamma & -\sin\alpha\cos\beta\sin\gamma+\cos\alpha\cos\gamma & \sin\alpha\sin\beta  \\ 
-\sin\beta\cos\gamma & \sin\beta\sin\gamma  & \cos\beta 
\end{matrix}\right],
\]
where $0\le\alpha,\gamma< 2\pi$ and $0\le\beta\le\pi$. These rotations do not commute, so the group is non-abelian.

@@

%% TODO: Separate section here on unitary representations? Existence of finite-dimensional

@@

For $SU(2)$ the generators taken from the fundamental representation $\mathbf 2$ were proportional to the Pauli matrices $X_i= \frac{1}{2}\sigma_i$, and the exponential map generating a group member $U$ is thus, as we alluded to earlier, $U=e^{i\alpha_iX^i}$, where $\alpha_i$ are the parameters. For $SO(3)$ the (Hermitian) generators taken from the fundamental representation $\mathbf 3$ are the matrices $Y_i=-iL_i$, where the $L_i$ were given in (\ref{eq:SO3_generators}), and the exponential map is $O=e^{i\alpha_iY^i}$.
Since $SU(2)$ and  $SO(3)$ have the same algebra for $X_i$ and $Y_i$, but the groups are {\it not} isomorphic,  in what sense, if any, does these exponential maps generate different elements? The answer to that is that $SU(2)$ is a {\bf double cover} of $SO(3)$, it has double the number of elements. There exists a two-to-one group homomorphism $f:SU(2)\to SO(3)$. This can be represented by $U(\alpha_1,\alpha_2,\alpha_3)\to O(\alpha_1,\alpha_2,\alpha_3)$. To see that this is two-to-one, assume that $\alpha_1=\alpha_2=0$. Then $O(0,0,\alpha_3)=O(0,0,\alpha_3+2\pi)$ since a rotation by an extra $2\pi$ does nothing. For $SU(2)$ we instead have
\[
U(0,0,\alpha_3)=\left[\begin{matrix} e^{i\frac{\alpha_3}{2}} & 0 \\ 0 & -e^{i\frac{\alpha_3}{2}} \end{matrix}\right] \quad \text{and}\quad U(0,0,\alpha_3+2\pi)=\left[\begin{matrix} e^{i\frac{\alpha_3}{2}+i\pi} & 0 \\ 0 & -e^{i\frac{\alpha_3}{2}+i\pi} \end{matrix}\right]=-U(0,0,\alpha_3)
\]
Thus the $SU(2)$ matrices has a period $4\pi$ versus the period $2\pi$ for $SO(3)$.\footnote{The argument can be made more formal and independent of the parameterisation, but is to extensive to repeat here.} Notice that this means that the three-dimensional representation of the Lie algebra coming from the generators of $SO(3)$ does not generate all the elements of $SU(2)$, while the two-dimensional one does.
% Better description of double cover in 3.2 https://www.weizmann.ac.il/particle/perez/Courses/QMII18/TA7.pdf

@@

We saw in Sec.~\ref{sec:unitary_groups} that we could construct unitary matrices from Hermitian ones by matrix exponentiation, but in fact for unitary groups the generators need to be Hermitian. To see this let a group member be $U=\exp(ia_iM^i)$. Then $U^\dagger = \exp(ia_iM^i)^\dagger=\exp(-ia_iM^{i\dagger})$,  but for a unitary matrix $U^\dagger = U^{-1}=\exp(-ia_iM^i)$, thus $M^{i\dagger}=M^i$ and the generators $M_i$ are Hermitian.

%{\bf MOVE DISCUSSION OF DYNKIN INDEX AND CASIMIR QUADRATIC HERE. SEE MASTER THESIS}
%Finally, we need two more calculation focused definitions for representations of Lie groups.
%Representations have some numerical properties that we are often interested in. 
%\df{For the representation $R$ the {\bf Casimir invariant} $C(R)$ is given by $C(R)\delta_{ij} = (T^a T^a)_{ij}$.}

%%%
\subsection{Adjoint representations}
\label{sec:adjoint_reps}
The definition of the structure constants in Sec.~\ref{sec:structure_constants} allows us to introduce another representation, the {\bf adjoint representation} of the algebra -- and by the exponential map the adjoint representation of the group -- where the representation of the algebra consists of the matrices $M_i$ with elements:
\[(M_i)_j^{~k} = -iC_{ij}^{~~k},\]
where $C_{ij}^{~~k},$ are the structure constants. If the dimension of the Lie algebra is $n$, then this is the number of generators, and it also means that the dimension of the matrices $M_i$ is $n\times n$. 
To prove that this is actually a representation, we need to check that the mapping from the elements of the algebra $X_i$ to the matrices $M_i$ satisfies the homomorphism criteria of the Lie algebra representations. From the Jacobi identity of the algebra it follows that
\begin{equation}
[M_i, M_j] = iC^{~~k}_{ij}M_k,
\label{eq:adjoint_rep_commutator}
\end{equation}
meaning that the matrices of the adjoint representation fulfils the same commutator relationship as the fundamental (generators). As a direct consequence the $M_i$ form a representation.

@@

\begin{Exercise}[]
Find the fundamental representation for $SO(3)$.
\end{Exercise}

\begin{Exercise}[difficulty={3},label=ex:SO2_diff_generators]
Find the differential generators for $SO(3)$ and their commutation relationships by studying how a function on $\mathbb R^3$ changes under $SO(3)$ transformations. Show that they satisfy the same commutators as the matrix generators.
\end{Exercise}

@@

\begin{Exercise}[title={Properties of matrix exponentiation},label={ex:expmap_prop},label=ex:expmapprop]
Prove the following useful properties of matrix exponentiation. $A$ and $B$ are matrices.
\begin{enumerate}
\item If $A$ and $B$ commute, $e^{A+B}=e^Ae^B$.
\item If $B$ has an inverse, $e^{BAB^{-1}}=Be^AB^{-1}$.
\item $e^{A^*}=(e^A)^*$
\item $e^{A^T}=(e^A)^T$
\item $e^{A^\dagger}=(e^A)^\dagger$
\item $e^{-A}=(e^A)^{-1}$
\item $\det e^A=e^{\tr{A}}$.
\end{enumerate}
\end{Exercise}

>>>>

We can construct unitary matrices from either Hermitian or anti-Hermitian matrices by using the {\bf matrix exponential}. This formally interprets the exponential series in terms of a real or complex valued $n\times n$ matrix $M$ as
\[ \exp(M)\equiv\sum_{n=0}^\infty \frac{M^n}{n!}=I+M+\frac{1}{2}M^2+\frac{1}{6}M^3+\ldots. \]
This series can be shown to always converge, so the series is well defined. Since this is a series of non-commuting objects we again have to be careful with using properties of the exponential from ordinary arithmetic. The following useful properties of the matrix exponential can be proven:
\begin{itemize}
\item[i)] $\exp(A^T)=\exp(A)^T$
\item[ii)] $\exp(A^*)=\exp(A)^*$
\item[iii)] If $B$ is an invertible matrix then $\exp(BAB^{-1})=B\exp(A)B^{-1}$
\item[iv)] If $[A,B]=0$ then $\exp(A)\exp(B)=\exp(A+B)$
\item[v)] $\det e^A = e^{\Tr A}$. 
\end{itemize}
% TODO: Discuss the use of Jordan canonical form in calculations here? Uses iii) to rewrite on Jordan canonical form $J=MAM^{-1}$ where $e^J$ is easy to evaluate because of the Jordan canonical form properties.
% Used in proofs?

@@

One particular explicit form, where the angles correspond to the three {\bf Euler angles} of rotation in three dimensions, $\alpha$, $\beta$ and $\gamma$, is
\begin{equation}
R(\alpha,\beta,\gamma)=\left[\begin{matrix} 
\cos\alpha\cos\beta\cos\gamma-\sin\alpha\sin\gamma & -\cos\alpha\cos\beta\sin\gamma-\sin\alpha\cos\gamma & \cos\alpha\sin\beta  \\ 
\sin\alpha\cos\beta\cos\gamma+\cos\alpha\sin\gamma & -\sin\alpha\cos\beta\sin\gamma+\cos\alpha\cos\gamma & \sin\alpha\sin\beta  \\ 
-\sin\beta\cos\gamma & \sin\beta\sin\gamma  & \cos\beta 
\end{matrix}\right],
\label{eq:R3_rotation_Euler}
\end{equation}
where $0\le\alpha,\gamma< 2\pi$ and $0\le\beta\le\pi$. These rotations do not commute, so the group is non-abelian.

@@

%% TODO: Separate section here on unitary representations? Relate to compact. Existence of finite-dimensional
%% TODO: Discuss the difference between active and passive view of transformation, e.g. Williams Sec. 1.2
% TODO: Linear representations as a special category

@@

For $SU(2)$ the generators taken from the fundamental representation $\mathbf 2$ were proportional to the Pauli matrices $X_i= \frac{1}{2}\sigma_i$, and the exponential map generating a group member $U$ is thus, as we alluded to earlier, $U=e^{i\alpha_iX^i}$, where $\alpha_i$ are the parameters. For $SO(3)$ the (Hermitian) generators taken from the fundamental representation $\mathbf 3$ are the matrices $Y_i=-iL_i$, where the $L_i$ were given in (\ref{eq:SO3_generators}), and the exponential map is $O=e^{i\alpha_iY^i}$.
Since $SU(2)$ and  $SO(3)$ have the same algebra for $X_i$ and $Y_i$, but the groups are {\it not} isomorphic,  in what sense, if any, does these exponential maps generate different elements? The answer to that is that $SU(2)$ is a {\bf double cover} of $SO(3)$, it has double the number of elements. There exists a two-to-one group homomorphism $f:SU(2)\to SO(3)$. This can be represented by $U(\alpha_1,\alpha_2,\alpha_3)\to O(\alpha_1,\alpha_2,\alpha_3)$. To see that this is two-to-one, assume that $\alpha_1=\alpha_2=0$. Then $O(0,0,\alpha_3)=O(0,0,\alpha_3+2\pi)$ since a rotation by an extra $2\pi$ does nothing. For $SU(2)$ we instead have
\[
U(0,0,\alpha_3)=\left[\begin{matrix} e^{i\frac{\alpha_3}{2}} & 0 \\ 0 & -e^{i\frac{\alpha_3}{2}} \end{matrix}\right] \quad \text{and}\quad U(0,0,\alpha_3+2\pi)=\left[\begin{matrix} e^{i\frac{\alpha_3}{2}+i\pi} & 0 \\ 0 & -e^{i\frac{\alpha_3}{2}+i\pi} \end{matrix}\right]=-U(0,0,\alpha_3)
\]
Thus the $SU(2)$ matrices has a period $4\pi$ versus the period $2\pi$ for $SO(3)$.\footnote{The argument can be made more formal and independent of the parameterisation, but is to extensive to repeat here.} Notice that this means that the three-dimensional representation of the Lie algebra coming from the generators of $SO(3)$ does not generate all the elements of $SU(2)$, while the two-dimensional one does.
% TODO: Better description of double cover in 3.2 https://www.weizmann.ac.il/particle/perez/Courses/QMII18/TA7.pdf

@@

We saw in Sec.~\ref{sec:unitary_groups} that we could construct unitary matrices from Hermitian ones by matrix exponentiation, but in fact for unitary groups the generators need to be Hermitian. To see this let a group member be $U=\exp(ia_iM^i)$. Then $U^\dagger = \exp(ia_iM^i)^\dagger=\exp(-ia_iM^{i\dagger})$,  but for a unitary matrix $U^\dagger = U^{-1}=\exp(-ia_iM^i)$, thus $M^{i\dagger}=M^i$ and the generators $M_i$ are Hermitian.
%TODO: Add tracelessness for $SU$. If $U=\exp(ia_iM^i)$ then $\det U = \exp{ia_i\Tr[M^i]}$ and if $\det U =1$ then we must have $\Tr M_i =0$ for all $i$.

%%%
\subsection{Adjoint representations}
\label{sec:adjoint_reps}
The definition of the structure constants in Sec.~\ref{sec:structure_constants} allows us to introduce another representation, the {\bf adjoint representation} of the algebra -- and by the exponential map the adjoint representation of the group -- where the representation of the algebra consists of the matrices $M_i$ with elements:
\[(M_i)_j^{~k} = -iC_{ij}^{~~k},\]
where $C_{ij}^{~~k},$ are the structure constants. If the dimension of the Lie algebra is $n$, then this is the number of generators, and it also means that the dimension of the matrices $M_i$ is $n\times n$. 
% TODO: Comment on when index position is important (not for semi-simple compact groups). Source?
To prove that this is actually a representation, we need to check that the mapping from the elements of the algebra $X_i$ to the matrices $M_i$ satisfies the homomorphism criteria of the Lie algebra representations. From the Jacobi identity of the algebra it follows that
\begin{equation}
[M_i, M_j] = iC^{~~k}_{ij}M_k,
\label{eq:adjoint_rep_commutator}
\end{equation}
meaning that the matrices of the adjoint representation fulfils the same commutator relationship as the fundamental (generators). As a direct consequence the $M_i$ form a representation.

@@

\begin{Exercise}[]
Find the fundamental matrix representation for $SO(3)$.
\end{Exercise}

\begin{Exercise}[difficulty={2},label=ex:SO2_diff_generators]
Find a differential representation of the generators for $SO(3)$ by studying how a vector in $\mathbb R^3$ changes under a parameterised $SO(3)$ transformation. {\it Hint}: Depending on the choice of parametrisation you may also need to use the commutator relationships between the generators.
\end{Exercise}
\begin{Answer}
Finding the differential generators for SO(3) begins with finding the composition function, or more specifically the elements of $\mathbf  x'=\mathbf f(\mathbf x, \boldsymbol \Lambda)$ where $\mathbf x$ is a vector in $\mathbb{R}^3$, $\boldsymbol \Lambda$ are the parameters of the transformation (the parameterisation of the manifold of the Lie group), and $\mathbf  x'$ the transformed vector. Using the parameterisation of rotations in $\mathbb{R}^3$ given in Eq.~(\ref{eq:R3_rotation_Euler}) ur composition function becomes 
$$f_i(\mathbf  x,\boldsymbol \Lambda) = R(\alpha,\beta,\gamma)_{ij} x_j,$$ 
with $\boldsymbol\Lambda= (\alpha,\beta,\gamma)$ containing the angles which we rotate around the different axis. Explicitly, the components thus become
\begin{align}
f_1(\mathbf  x,\boldsymbol \Lambda) &= (\cos\alpha\cos\beta\cos\gamma - \sin\alpha\sin\gamma)x - (\cos\alpha\cos\beta\sin\gamma+\sin\theta\cos\gamma)y + \cos\alpha\sin\beta z \\
f_2(\mathbf  x,\boldsymbol\Lambda) &= (\sin\alpha\cos\beta\cos\gamma + \cos\alpha\sin\gamma)x + (\cos\alpha\cos\gamma-\sin\alpha\cos\beta\sin\gamma)y + \sin\alpha\sin\beta z\\
f_3(\mathbf  x,\boldsymbol\Lambda) &= -\sin\beta\cos\gamma x + \sin\beta\sin\gamma y + \cos\beta z
\end{align}

The next step is finding the generators $X^j$ from the definition
$$iX^j = \left.\frac{\partial f_i}{\partial\Lambda_j}\frac{\partial}{\partial x_i}\right|_{\boldsymbol\Lambda = 0}.$$
To simplify notation we abbreviate $\frac{\partial}{\partial x} =\partial_x$. Doing the calculation for all three parameters of the manifold we get three generators, as follows. 
$$iX^1 = x\partial_y-y\partial_x \Rightarrow L_z\equiv X^1 = -i(x\partial_y-y\partial_x),$$
$$iX^2 = z\partial_x-x\partial_z \Rightarrow L_y\equiv X^2 = -i(z\partial_x-x\partial_z),$$
$$iX^3 = x\partial_y-y\partial_x \Rightarrow L_z\equiv X^3 = -i(x\partial_y-y\partial_x).$$
The generators we found are proportional to (two of) the angular momentum operators in quantum mechanics.
Unfortunately, only two of the generators are unique. This is due to the choices made in the parametrisation of the manifold, and will depend on which parametrisation you start with. To find the last generator of our group, all we need to do here is to use the commutator relations between the generators. We start with
\begin{eqnarray*}
\left[L_y,L_z\right]f &=& -(z\partial_x-x\partial_z)(x\partial_y-y\partial_x)f + (x\partial_y-y\partial_x)(z\partial_x-x\partial_z)f \\
&=& -z\partial_y f -zx\partial_x\partial_y f + zy\partial_x^2 f+x^2\partial_z\partial_y f - xy\partial_z\partial_x f \\
&&+ xz\partial_y\partial_x f-x^2\partial_y\partial_z f-yz\partial_x^2 f+y\partial_zf+yx\partial_x\partial_z f \\
&=& i(-i)(y\partial_z-z\partial_y)f,
\end{eqnarray*}
where we have use that the partial derivatives commute. Since we know from our commutators that $\left[L_y,L_z\right]=iL_x$, our last generator becomes $$L_x=-i(y\partial z-z\partial y).$$
 %The last two commutators are $$\left[L_x,L_y\right] = -y\partial_x+x\partial_y = i(-i)(x\partial_y-y\partial_x) = iL_z.$$
%$$\left[L_z,L_x\right] = -x\partial_z+z\partial_x = i(-i)(z\partial_x-x\partial_z) = iL_y.$$
\end{Answer}

@@

\begin{Exercise}[title={Properties of matrix exponentiation},label={ex:expmap_prop}]
Prove the following useful properties of matrix exponentiation. $A$ and $B$ are matrices.
\begin{enumerate}
\item If $A$ and $B$ commute, $e^{A+B}=e^Ae^B$.
\item If $B$ has an inverse, $e^{BAB^{-1}}=Be^AB^{-1}$.
\item $e^{A^*}=(e^A)^*$
\item $e^{A^T}=(e^A)^T$
\item $e^{A^\dagger}=(e^A)^\dagger$
\item $e^{-A}=(e^A)^{-1}$
\item $\det e^A=e^{\tr{A}}$.
\end{enumerate}
\end{Exercise}
